Ab Initio Effective One-Electron Potential Operators. II. Applications for Exchange-Repulsion Energy in Effective Fragment Potentials

### 2020-02-21 papers/2_oep_technique.CT/paper-r3.tex

%
\begin{figure*}[h]
\includegraphics[width=0.9\textwidth]{fig-2.pdf}
\caption{\label{f:fig-2} {\bf Accuracy of the ERI elimination technique.}
(a) NCB31 
database,\cite{Zhao.Schultz.Truhlar.JCTC.2006,
Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005} 
and
(b) BBI subset\cite{Burns.Faver.Zheng.Marshall.Smith.Vanommeslaeghe.MacKerell.Merz.Sherrill.JCP.2017} 
from the BioFragment Database.
The EDF-1 scheme with the aug-cc-pVDZ-jkfit auxiliary basis set
for the OEP calculations.
} 
\end{figure*}
%

@@

The utmost goal of this work is to reduce the computational cost of
the CT/HF0 energy evaluation in the calculations involving effective fragment
potentials, no longer making it the bottleneck of the EFP2\hyp{}based interaction
energy calculations. In Table~\ref{t:oep-costs} estimation of the
computational cost of the EFP2 and OEP models is shown.
It is apparent that EFP2 has much more quantities to compute on the fly
as compared to the OEP method (`Calculables' in the table).
Clearly,
evaluation of the EFP2 CT expression from Eq.~\eqref{e:u-efp2-ct} 
requires quite a number of different types of OEI's. According to our estimations
that assume sequential (two\hyp{}step) two\hyp{}index AO\hyp{}MO transformations of OEI matrices
and large AO basis sets,
the computational cost
has an order of $2p^3(s+t) + 3vop^2$, where the
$o$ and $p$ denote the number of occupied orbitals and the number of atomic basis functions,
respectively. 
$s$, $t$ and $v$ are the relative costs of evaluation of the
overlap, kinetic energy, and multipole potential OEI's, respectively, with the latter being
most expensive but
%type of OEI's, i.e., the potential and potential gradient(s)
%integrals contracted with the DMTP's, 
%that are 
necessary to compute ${\bf V}^{\rm EF}$ matrices
from Eq.~\eqref{e:v-efp2-eff}. 
%
{
\renewcommand{\arraystretch}{1.4}
\begin{table}[t]
\caption[Minimal uncontracted auxiliary basis set optimized for OEP\hyp{}based CT/HF0 calculations]
{{\bf Minimal uncontracted auxiliary basis set optimized for OEP\hyp{}based CT/HF0 calculations\footnotemark[1]}
}
\label{t:3}
\begin{ruledtabular}
\begin{tabular}{lldcd}
 AO's &  & \multicolumn{3}{c}{Exponents (a.u.)}\\
      &  & \text{Methanol}         && \text{Water}           \\
\cline{1-2}
\cline{3-5}
 H    &1s&   7.7369294534   &&   29.5837988322 \\
 C    &1s& 880.3654511045   &&                 \\
      &2s& 123.8276940826   &&                 \\
      &2p&  11.6748155072   &&                 \\
 O    &1s& 902.9798919021   && 1030.5721050297 \\
      &2s& 203.7584408153   &&  142.1022267153 \\
      &2p&  20.2442429692   &&   40.9175702474 \\
\end{tabular}
\end{ruledtabular}
%
\footnotetext[1]{Fitting performed according to Appendix~\eqref{a:auxiliary-basis} assuming 6-311++G(d,p) 
and aug-cc-pVDZ-jkfit primary and test basis sets, respectively.}
%
\end{table}
}
%
On the contrary, 
OEP\hyp{}based expression
from Eq.~\eqref{e:ect-oep.final} requires only overlap OEI's that are the least expensive, 
and has the cost of approximate magnitude of $2sop^2$ for relatively small auxiliary basis sets.
%where 
%$Q$ and $N$ denote the number of auxiliary basis functions and the number of atoms,
%respectively. 
Note also that, among the calculables that are needed on the fly in each
OEP\hyp{}based CT energy evaluation, are the auxiliary vectors and matrices 
from Eqs.~\eqref{e:ct-murell-etal.vin.oep.aux-var.u}
and \eqref{e:ct-murell-etal.vin.oep.aux-var.w},
the cost of which is negligible. The amount of effective fragment parameters,
that need to be superimposed during the calculations by applying rotation
of orbitals and basis functions (`Superimposable parameters')
is rather the same in EFP2 and OEP models. 
This includes the LCAO\hyp{}MO coefficients in canonical basis,
and the primary basis set, with an addition of the auxiliary basis set
for the OEP model. Therefore, the cost of parameter superimposition should
not be significantly larger as in the EFP2 formulation, provided sufficiently 
small auxiliary basis set is used.
For example,
assuming a water dimer system and 6-311++G(d,p) primary and minimal auxiliary basis set
with $s=t=v\approx1$,
the OEP method is predicted to be roughly 12--16 times faster than EFP2 method. In practice,
the parameters $s$, $t$ and $v$ will have larger values, especially $v$.

>>>>

%
\begin{figure*}[h]
\includegraphics[width=0.9\textwidth]{fig-2.pdf}
\caption{\label{f:fig-2} {\bf Accuracy of the ERI elimination technique.}
(a) NCB31 
database,\cite{Zhao.Schultz.Truhlar.JCTC.2006,
Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005} 
and
(b) BBI subset\cite{Burns.Faver.Zheng.Marshall.Smith.Vanommeslaeghe.MacKerell.Merz.Sherrill.JCP.2017} 
from the BioFragment Database.
For the OEP calculations, the EDF-1 scheme with the aug-cc-pVDZ-jkfit auxiliary basis set
was used.
} 
\end{figure*}
%

@@

The utmost goal of this work is to reduce the computational cost of
the CT/HF0 energy evaluation in the calculations involving effective fragment
potentials, no longer making it the bottleneck of the EFP2\hyp{}based interaction
energy calculations. In Table~\ref{t:oep-costs} estimation of the
computational cost of the EFP2 and OEP models is shown.
It is apparent that EFP2 has much more quantities to compute on the fly
as compared to the OEP method (`Calculables' in the table).
Clearly,
evaluation of the EFP2 CT expression from Eq.~\eqref{e:u-efp2-ct} 
requires quite a number of different types of OEI's. According to our estimations
that assume sequential (two\hyp{}step) two\hyp{}index AO\hyp{}MO transformations of OEI matrices
and large AO basis sets,
the computational cost
has an order of $2p^3(s+t) + 3vop^2$, where the
$o$ and $p$ denote the number of occupied orbitals and the number of atomic basis functions,
respectively. 
$s$, $t$ and $v$ are the relative costs of evaluation of the
overlap, kinetic energy, and multipole potential OEI's, respectively, with the latter being
most expensive but
%type of OEI's, i.e., the potential and potential gradient(s)
%integrals contracted with the DMTP's, 
%that are 
necessary to compute ${\bf V}^{\rm EF}$ matrices
from Eq.~\eqref{e:v-efp2-eff}. 
%
{
\renewcommand{\arraystretch}{1.4}
\begin{table}[t]
\caption[Minimal uncontracted auxiliary basis set optimized for OEP\hyp{}based CT/HF0 calculations]
{{\bf Minimal uncontracted auxiliary basis set optimized for OEP\hyp{}based CT/HF0 calculations\footnotemark[1]}
}
\label{t:3}
\begin{ruledtabular}
\begin{tabular}{lldcd}
 AO's &  & \multicolumn{3}{c}{Exponents (a.u.)}\\
      &  & \text{Methanol}         && \text{Water}           \\
\cline{1-2}
\cline{3-5}
 H    &1s&   7.7369294534   &&   29.5837988322 \\
 C    &1s& 880.3654511045   &&                 \\
      &2s& 123.8276940826   &&                 \\
      &2p&  11.6748155072   &&                 \\
 O    &1s& 902.9798919021   && 1030.5721050297 \\
      &2s& 203.7584408153   &&  142.1022267153 \\
      &2p&  20.2442429692   &&   40.9175702474 \\
\end{tabular}
\end{ruledtabular}
%
\footnotetext[1]{Fitting performed according to Appendix~\eqref{a:auxiliary-basis} assuming 6-311++G(d,p) 
and aug-cc-pVDZ-jkfit primary and test basis sets, respectively.}
%
\end{table}
}
%
\\ On the contrary, 
OEP\hyp{}based expression
from Eq.~\eqref{e:ect-oep.final} requires only overlap OEI's that are the least expensive, 
and has the cost of approximate magnitude of $2sop^2$ for relatively small auxiliary basis sets.
%where 
%$Q$ and $N$ denote the number of auxiliary basis functions and the number of atoms,
%respectively. 
Note also that, among the calculables that are needed on the fly in each
OEP\hyp{}based CT energy evaluation, are the auxiliary vectors and matrices 
from Eqs.~\eqref{e:ct-murell-etal.vin.oep.aux-var.u}
and \eqref{e:ct-murell-etal.vin.oep.aux-var.w},
the cost of which is negligible. The amount of effective fragment parameters,
that need to be superimposed during the calculations by applying rotation
of orbitals and basis functions (`Superimposable parameters')
is rather the same in EFP2 and OEP models. 
This includes the LCAO\hyp{}MO coefficients in canonical basis,
and the primary basis set, with an addition of the auxiliary basis set
for the OEP model. Therefore, the cost of parameter superimposition should
not be significantly larger as in the EFP2 formulation, provided sufficiently 
small auxiliary basis set is used.
For example,
assuming a water dimer system and 6-311++G(d,p) primary and minimal auxiliary basis set
with $s=t=v\approx1$,
the OEP method is predicted to be roughly 12--16 times faster than EFP2 method. In practice,
the parameters $s$, $t$ and $v$ will have larger values, especially $v$.

### 2020-02-21 papers/2_oep_technique.CT/supplementary-r2.tex

\input{header.tex}
\begin{document}
\preprint{AIP/123-OEP}

@@

%
\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{fig-s1.pdf}
\caption{\label{f:fig-s1} {\bf Asymptotic dependence of the charge transfer energy
in the CT/HF0 formulation for water-methanol complex.} 
Water molecule has been translated by $R$ from the starting geometry
along the vector specified in the inset picture.
The counterpoise\hyp{}corrected total interaction energy
is also shown for comparison in light grey color in this figure.
All data were obtained at HF/6-311++G(2df,2pd) level of theory
with the single GDF scheme and aug-cc-pVTZ-jkfit auxiliary basis set.
} 
\end{figure}
%

@@

>>>>

\documentclass[aip,jcp,amsmath,amssymb,preprint,floatfix]{revtex4-1}
\input{header.tex}
\begin{document}
\preprint{AIP/123-OEP}

% numbering for Figures in SI
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

@@

%
\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{fig-s1.pdf}
\caption{\label{f:fig-s1} {\bf Asymptotic dependence of the charge transfer energy
in the CT/HF0 formulation for water-methanol complex.} 
Water molecule has been translated by $R$ from the starting geometry
along the vector specified in the inset picture.
The counterpoise\hyp{}corrected total interaction energy
is also shown for comparison in light grey color in this figure.
All data were obtained at HF/6-311++G(d,p) level of theory
with the EDF-1 scheme and aug-cc-pVQZ-jkfit auxiliary basis set.
} 
\end{figure}
%

@@

In Figure~\ref{f:fig-s2} interaction energy components
obtained by using the code developed in this work are
compared to the ones obtained by using the EFP2 calculation routines
in GAMESS US quantum chemistry package\cite{GAMESS.JCC.1993} (version: Sept 30, 2017 R2 Public Release). 
For this, EFP2 parameters
were generated by running the \texttt{MAKEFP} routine
with \texttt{CTVVO} option switched to \texttt{.FALSE.} which corresponds
to using all the canonical molecular orbitals from HF calculations
instead of valence virtual orbitals.
Generally, Coulombic energies (Figure~S2(a)) 
are in very good agreement with a few exeptions in the PPS5 set,
for which the differences are due to the inaccuracies in the DMTP
expansion used in the EFP2 model -- note that the Coulombic component in the DDS/HF method
is free from multipole approximation and is equivalent to the first\hyp{}order SAPT Coulombic energy.
Exchange\hyp{}repulsion and induction energies (Figures~S2(b) and S2(c), respectively)
are in very good agreement. CT energies (Figure~S2(d)) 
evaluated with CAMM up to quadrupoles for the potential energy matrix elements (black filled circles)
are generally also in decent quantitative agreement,
except for NH$_3$--FCl, (HCOOH)$_2$ and (HCONH$_2$)$_2$ 
%and parallel\hyp{}displaced benzene dimer
where the differences
are between 2--4 kcal/mol. Only in the case of the latter three systems,
adding distributed octupoles (blue crosses in Figure~S2(d)) 
lowers the CT energy values by around 1 kcal/mol,
with negligible changes for the rest.

%
\begin{figure}[h]
\includegraphics[width=0.8\textwidth]{fig-s2.pdf}
\caption{\label{f:fig-s2} {\bf Validation of implementation of the interaction energy components
with respect to the EFP2 GAMESS US code.} 
Results obtained for the NCB31 database set assuming 6-311++G(d,p) primary basis set. 
(a) - Coulombic, (b) - exchange-repulsion, (c) - induction, and (d) - charge-transfer
interaction energy. In (d), black circles and blue crosses correspond to the
CT EFP2 energies with CAMM up to distributed quadrupoles and octupoles, respectively.
For more details see the main text (Section III, \emph{Calculation Details}).
} 
\end{figure}
%

### 2020-02-21 papers/3_oep_technique.Pauli/paper-r1.tex

\begin{abstract}
Recently, it was shown that
the effective one\hyp{}electron potentials (OEP) method
of effective elimination of the electron repulsion integrals
can be used to lower the computational cost
of evaluation of the charge transfer interaction energy
within the effective fragment potential method (EFP2)
by the factor of 20, making it no longer a bottleneck
in EFP2\hyp{}based simulations of complex systems.
Here, OEP technique 
is used to enhance computational efficiency in evaluating the
exchange\hyp{}repulsion EFP2 interaction energy
by incorporating the kinetic energy operator in the OEP
via the extended density fitting in incomplete auxiliary basis.
It is shown that the computational cost can be reduced by a factor
of 1.5 and kinetic energy integrals are no longer necessary to evaluate.
\end{abstract}

@@

In this work, an alternative formulation for the first\hyp{}order
term is proposed based on the effective one\hyp{}electron potential (OEP)
operator technique developed in the preceeding contribution.~\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019}
It is shown that calculation of kinetic energy integrals can be avoided
by introducing OEP's.

@@

We implemented EXR energy models from Eq.~\eqref{e:exr-efp2}
(EFP2), Eq.~\eqref{e:rep.murrell-etal.S1.oep} (OEP), as well as 
the approximate intermolecular perturbation theory with exchange of 
Murrell et~al.\cite{Murrell.Randic.Williams.Longuet-Higgins.ProcRSocLondA.1965},
%from Eq.~\eqref{e:rep.murrell-etal.S1} 
the exact EXR energy of Hayes \& Stone\cite{Hayes.Stone.MolPhys.1984}
and the density decomposition scheme 
at the Hartree\hyp{}Fock level of theory (here referred to as the DDS/HF) 
developed by Mandado and Hermida\hyp{}Ram{\'o}n,\cite{Mandado.Hermida-Ramon.JCTC.2011} 
for validation of the EDF technique.
When implemented in the dimer\hyp{}centered basis set\cite{Chalasinski.Gutowski.MolPhys.1985} (DCBS),
the DDS/HF model is
equivalent to the exact EXR energy by Hayes \& Stone and gives identical results,
which is the case in this work. Using DCBS also allows one to eliminate the basis set
superimposition error (DCBS).
To localize molecular orbitals, procedure that is necessary for the EFP2 calculations,
the Boys method\cite{Boys.RevModPhys.1960} was used. Note here that, since
Murrell et~al.'s theory is invariant with respect to unitary transformation
of molecular orbitals, the OEP\hyp{}based expression 
in Eq.~\eqref{e:rep.murrell-etal.S1.oep} is also invariant and does not require
orbital localization. 
%in which
%
%\begin{equation}\label{e:dds-exr}
% \Delta E^{\rm Ex-Rep}_{\rm DDS/HF} = 
%\end{equation}
%
%
\begin{figure*}[t]
\includegraphics[width=0.8\textwidth]{fig-2.pdf}
\caption{\label{f:fig-2} {\bf Accuracy of the OEP and EFP2 models of exchange\hyp{}repulsion energy
across various bi\hyp{}molecular systems.} 
(a) NCB31 
database,\cite{Zhao.Schultz.Truhlar.JCTC.2006,
Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005} 
of non\hyp{}covalent interactions
and
(b) BBI subset\cite{Burns.Faver.Zheng.Marshall.Smith.Vanommeslaeghe.MacKerell.Merz.Sherrill.JCP.2017} 
of backbone\hyp{}backbone interactions in proteins from the BioFragment Database.
For the OEP calculations, the EDF-1 scheme with the aug-cc-pVDZ-jkfit auxiliary basis set
was used.
} 
\end{figure*}
%

Three complexes: 
(H$_2$O)$_2$, 
H$_2$O--HOCH$_3$, and
(DMSO)$_2$,
were chosen as model systems to analyze the asymptotic dependence 
of EXR energy, which is shown in Figure~\ref{f:fig-1}. 
The reference (zero\hyp{}displacement) geometries
were obtained as described in Ref.\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019} 
and the structures  
along with the translation direction are depicted
in the insets in Figure~\ref{f:fig-1}.
Energy\hyp{}optimizations were performed at the HF/6-31+G(d,p) level,
as implemented in 
the {\sc Gaussian16} quantum chemistry program package.\cite{Gaussian16}
OEP and EFP2 models correctly describe the EXR energy at all separations
for the studied model systems with very similar accuracy. 
However,
contrary to our previous application of the OEP technique\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019},
where approximately 20\hyp{}fold speedups as compared to the EFP2 model were achieved
for the evaluation of the charge\hyp{}transfer energy,
the computational cost of the OEP EXR method is roughly 2 times higher than the EFP2 model
when the usual density fitting auxiliary basis sets such as aug-cc-pVDZ-jkfit or aug-cc-pVDZ-ri are used
(Table~\ref{t:tab-1}).
However, application of our small auxiliary basis sets for H$_2$O, CH$_3$OH and DMSO, 
%developed
developed via the basis set optimization~\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019}
in conjunction with the basin hopping global optimization algorithm  %cite
as implemented in the SciPy Python library
(see Supplementary Information),
%to minimize the computational cost of evaluating the summation terms
%from Eq.~\eqref{e:rep.murrell-etal.S1.oep},
reduces costs appreciably and makes the OEP method on average 1.5 times faster than EFP2.
For instance, calculation of the EXR energy in the DMSO dimer by using 6-31++G(d,p) primary
basis set and 
1.2 GHz AMD EPYC\texttrademark{} 7301 16-Core Processor
requires about 4.6~ms and 3.0~ms when EFP2 and OEP model is used, respectively.
%Note that the benchmark Hayes \& Stone's calculations predict slightly larger repulsion energies

To investigate the accuracy of the OEP\hyp{}based repulsion
term across
%perform statistical error analysis for 
a variety of
interacting systems, range of
bi\hyp{}molecular complexes from the non\hyp{}covalent
interactions database NCB31 developed by the Truhlar's 
group,\cite{Zhao.Schultz.Truhlar.JCTC.2006,
Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005}
as well as the BioFragment Database subset BBI for backbone\hyp{}backbone
interactions in proteins of Sherrill group,\cite{Burns.Faver.Zheng.Marshall.Smith.Vanommeslaeghe.MacKerell.Merz.Sherrill.JCP.2017} 
as implemented in the {\sc Psi4}
program,\cite{Psi4.JCTC.2017}
were utilized, and the
EXR energies were compared with the reference DDS/HF results (Figure~\ref{f:fig-2}).
On average, OEP and EXR tend to underestimate the EXR energy by 0.7\% and 7.6\%
in the NCB31/6-311++G(d,p) data set, and 6.0\% and 15\% in the BBI/6-31+G(d,p) data set,
respectively.
This is mostly due to the lack of BSSE, that is accounted for
only in the reference calculations with the DDS/HF model, and the effect is usually larger
when smaller primary basis sets are used.
The OEP model is of slightly better accuracy as the EFP2 model
in systems sudied in this work, with 
root mean square errors of OEP and EFP2 model being
2.0 kcal/mol and 2.6 kcal/mol for NCB31/6-311++G(d,p) data set, 
as well as
0.5 kcal/mol and 1.2 kcal/mol for BBI/6-31+G(d,p) data set, respectively.
It is an accidental fruitful cancellation of errors because
the first\hyp{}order repulsive terms have much less differences with respect to each other
(RMSE is 0.7\% only in all data sets).
Nevertheless, together with the performance data of both models
from Table~\ref{t:tab-1}, OEP model performs better than EFP2 as long as the minimal
auxiliary basis sets are used.

>>>>

\begin{abstract}
In Paper~I, it was shown that
the effective one\hyp{}electron potentials (OEP) method
of effective elimination of the electron repulsion integrals
can be used to lower the computational cost
of evaluation of the charge transfer interaction energy
within the effective fragment potential method (EFP2)
by the factor of 20, making it no longer a bottleneck
in EFP2\hyp{}based simulations of complex systems.
Here, OEP technique 
is used to enhance computational efficiency in evaluating the
exchange\hyp{}repulsion EFP2 interaction energy
by re\hyp{}defining the first\hyp{}order repulsive term of 
Murrell~et al. through
the extended density fitting in incomplete auxiliary basis.
It is shown that the computational cost can be reduced by a factor
of 1.5 as compared to the original EFP2 formulation
and kinetic energy integrals are no longer necessary to evaluate.
\end{abstract}

@@

In this work, an alternative formulation for the first\hyp{}order
term is proposed based on the effective one\hyp{}electron potential (OEP)
operator technique developed in the preceeding contribution.~\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019}
It is shown that calculation of kinetic energy integrals can be avoided
by introducing OEP's leading to enhancepent of computational efficiency.

@@

In our in\hyp{}house plugin to the {\sc Psi4}
quantum chemistry
program,\cite{Psi4.JCTC.2017}
we implemented EXR energy models from Eq.~\eqref{e:exr-efp2}
(EFP2), Eq.~\eqref{e:rep.murrell-etal.S1.oep} (OEP), as well as 
the approximate intermolecular perturbation theory with exchange of 
Murrell et~al.\cite{Murrell.Randic.Williams.Longuet-Higgins.ProcRSocLondA.1965},
%from Eq.~\eqref{e:rep.murrell-etal.S1} 
the exact EXR energy of Hayes \& Stone\cite{Hayes.Stone.MolPhys.1984},
here referred to as the HS model.
The latter was
implemented in the dimer\hyp{}centered basis set\cite{Chalasinski.Gutowski.MolPhys.1985} (DCBS),
to eliminate the basis set
superimposition error (DCBS) in the benchmark calculations.
%and the density decomposition scheme 
%at the Hartree\hyp{}Fock level of theory (here referred to as the DDS/HF) 
%developed by Mandado and Hermida\hyp{}Ram{\'o}n,\cite{Mandado.Hermida-Ramon.JCTC.2011} 
%for validation of the EDF technique.
%When implemented in the dimer\hyp{}centered basis set\cite{Chalasinski.Gutowski.MolPhys.1985} (DCBS),
%the DDS/HF model is
%equivalent to the exact EXR energy by Hayes \& Stone and gives identical results,
%which is the case in this work. 
%Using DCBS also allows one to eliminate the basis set
%superimposition error (DCBS).
To localize molecular orbitals, procedure that is necessary for the EFP2 calculations,
the Boys method\cite{Boys.RevModPhys.1960} was used. Note here that, since
Murrell et~al.'s theory is invariant with respect to unitary transformation
of molecular orbitals, the OEP\hyp{}based expression 
in Eq.~\eqref{e:rep.murrell-etal.S1.oep} is also invariant and does not require
orbital localization. 
%in which
%
%\begin{equation}\label{e:dds-exr}
% \Delta E^{\rm Ex-Rep}_{\rm DDS/HF} = 
%\end{equation}
%
%
\begin{figure*}[t]
\includegraphics[width=0.8\textwidth]{fig-2.pdf}
\caption{\label{f:fig-2} {\bf Accuracy of the OEP and EFP2 models of exchange\hyp{}repulsion energy
across various bi\hyp{}molecular systems.} 
(a) NCB31 
database,\cite{Zhao.Schultz.Truhlar.JCTC.2006,
Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005} 
of non\hyp{}covalent interactions
and
(b) BBI subset\cite{Burns.Faver.Zheng.Marshall.Smith.Vanommeslaeghe.MacKerell.Merz.Sherrill.JCP.2017} 
of backbone\hyp{}backbone interactions in proteins from the BioFragment Database.
For the OEP calculations, the EDF-1 scheme with the aug-cc-pVDZ-jkfit auxiliary basis set
was used.
} 
\end{figure*}
%

Three complexes: 
(H$_2$O)$_2$, 
H$_2$O--HOCH$_3$, and
(DMSO)$_2$,
were chosen as model systems to analyze the asymptotic dependence 
of EXR energy, which is shown in Figure~\ref{f:fig-1}. 
The reference (zero\hyp{}displacement) geometries
were obtained as described in Ref.\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019} 
and the structures  
along with the translation direction are depicted
in the insets in Figure~\ref{f:fig-1}.
Energy\hyp{}optimizations were performed at the HF/6-31+G(d,p) level,
as implemented in 
the {\sc Gaussian16} quantum chemistry program package.\cite{Gaussian16}
OEP and EFP2 models correctly describe the EXR energy at all separations
for the studied model systems with very similar accuracy. 
However,
contrary to our previous application of the OEP technique\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019},
where approximately 20\hyp{}fold speedups as compared to the EFP2 model were achieved
for the evaluation of the charge\hyp{}transfer energy,
the computational cost of the OEP EXR method is roughly 2 times higher than the EFP2 model
when the usual density fitting auxiliary basis sets such as aug-cc-pVDZ-jkfit or aug-cc-pVDZ-ri are used
(Table~\ref{t:tab-1}).
Computational cost of $\Delta E^{\rm Rep}(S^{-1})$ in EFP2 model
is approximately $s(on^2 + o^2n) + t(on^2 + o^2n) + 2o^3$, whereas in OEP model
is $s(on^2 + o^2n) + 2ao^2$, both of which have comparable magnitudes. 
Here, $s$ and $t$ denote the relative cost of evaluation of the overlap
and kinetic energy one electron integrals (OEI's), whereas $o$, $n$ and $a$ is
the number of LMO's, primary AO's and auxiliary AO's, respectively.
It is clear that the critical parameter is the number of auxiliary basis functions $a$,
which should be comparable with the number of occupied orbitals $o$ 
as $t$ is usually only 2--3 times larger than $s$.
We developed small auxiliary basis sets for H$_2$O, CH$_3$OH and DMSO, 
%developed
via the basis set optimization method described in Appendix~A to Paper I
in conjunction with the basin hopping global optimization algorithm\cite{Wales.EnergyLandscapes.2003,
Wales.Scheraga.Science.1999,
Wales.Doye.JPCA.1997,Li.Scheraga.PNAS.1987}
as implemented in the SciPy Python library\cite{SciPy.2019}
(see Supplementary Information).
%to minimize the computational cost of evaluating the summation terms
%from Eq.~\eqref{e:rep.murrell-etal.S1.oep},
Applying such basis sets
reduces costs appreciably and makes the OEP method on average 1.5 times faster than EFP2.
For instance, calculation of the EXR energy in the DMSO dimer by using 6-31++G(d,p) primary
basis set and 
1.2~GHz AMD EPYC\texttrademark{} 7301 16-Core Processor
requires about 4.6~ms and 3.0~ms when EFP2 and OEP model is used, respectively.
%Note that the benchmark Hayes \& Stone's calculations predict slightly larger repulsion energies

To investigate the accuracy of the OEP\hyp{}based repulsion
term across
%perform statistical error analysis for 
a variety of
interacting systems, range of
bi\hyp{}molecular complexes from the non\hyp{}covalent
interactions database NCB31 developed by the Truhlar's 
group,\cite{Zhao.Schultz.Truhlar.JCTC.2006,
Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005}
as well as the BioFragment Database subset BBI for backbone\hyp{}backbone
interactions in proteins of Sherrill group,\cite{Burns.Faver.Zheng.Marshall.Smith.Vanommeslaeghe.MacKerell.Merz.Sherrill.JCP.2017} 
as implemented in the {\sc Psi4}
program,\cite{Psi4.JCTC.2017}
were utilized, and the
EXR energies were compared with the reference HS results (Figure~\ref{f:fig-2}).
On average, OEP and EXR tend to underestimate the EXR energy by 0.7\% and 7.6\%
in the NCB31/6-311++G(d,p) data set, and 6.0\% and 15\% in the BBI/6-31+G(d,p) data set,
respectively.
This is mostly due to the lack of BSSE, that is accounted for
only in the reference calculations with the HS model, and the effect is usually larger
when smaller primary basis sets are used.
The OEP model is of slightly better accuracy as the EFP2 model
in systems sudied in this work, with 
root mean square errors of OEP and EFP2 model being
2.0 kcal/mol and 2.6 kcal/mol for NCB31/6-311++G(d,p) data set, 
as well as
0.5 kcal/mol and 1.2 kcal/mol for BBI/6-31+G(d,p) data set, respectively.
It is an accidental fruitful cancellation of errors because
the EFP2 and OEP estimations of 
$\Delta E^{\rm Rep}(S^{-1})$ have much less differences with respect to each other
(RMSE between EFP2 and OEP is only 0.7\% in all data sets).
Nevertheless, together with the performance data of both models
from Table~\ref{t:tab-1}, OEP model performs better than EFP2 as long as the minimal
auxiliary basis sets are used.
